---
title: "10X Seurat QC Demo"
author:
  - name: "Aishwarya Sundaresan (CCBB)"
date: "`r Sys.Date()`"  
output: html_document
---
  <style>
  pre {
    overflow-x: auto;
  }
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(fig.width=15)
```
```{r Loading libraries,warning=FALSE,message=FALSE,echo=FALSE}
library(Seurat)
library(hdf5r)
library(readr)
library(dplyr)
library(tidyverse)
library(Matrix)
library(patchwork)
library(clustree)
library(ggplot2)
library(sctransform)
library(future)
options(future.globals.maxSize = 10000 * 1024^2)
plan("multisession", workers = 4)
```
Please contact [CCBB](https://www.scripps.edu/science-and-medicine/cores-and-services/center-for-computational-biology-bioinformatics/) for analysis! \

This document provides QC metrics and associated plots to visually explore the quality of the single cell data. \

Single cell data is initially filtered in order to exclude genes that are expressed in fewer than 100 cells,and to exclude cells that contain fewer than 100 expressed genes. \

```{r Reading data,warning=FALSE,message=FALSE,echo=FALSE}
#Read in the  count matrix hdf5 generated by the cellranger pipeline
data.10x_file_names<-list.files(path ="./",pattern = ".h5")
data.10x = list(); # first declare an empty list in which to hold the count matrices
for( i in 1:length(data.10x_file_names) ){
  data.10x[[i]] <- Read10X_h5(file=paste("./",data.10x_file_names[i],sep="/"),use.names = T, unique.features = T)
}

#Create a vector of convenient sample names
samples<-sapply(strsplit(data.10x_file_names, "_filtered_feature_bc_matrix"), "[", 1)

#Create a variable called outdir to specify your output directory
scrna_seurat_qc_outdir = "scrna_seurat_qc"
if (!dir.exists(scrna_seurat_qc_outdir)) {dir.create(scrna_seurat_qc_outdir)}

#Convert each count matrix to a Seurat object
#This simultaneously performs some initial filtering in order to exclude genes that are expressed in fewer than 100 cells,
#and to exclude cells that contain fewer than 100 expressed genes. 
#At this step, we also create a “group” identity for each cel and and prepend the appropriate data set name to each cell barcode. 
#For example, if a barcode from data set “B” is originally AATCTATCTCTC, it will now be B_AATCTATCTCTC. 

scrna.list = list(); # First create an empty list to hold the Seurat objects

for (i in 1:length(data.10x)) {
  scrna.list[[i]] = CreateSeuratObject(counts = data.10x[[i]], min.cells=100, min.features=100, project=samples[i]);
  scrna.list[[i]][["Group"]] = samples[i];
  scrna.list[[i]] <- RenameCells(scrna.list[[i]],add.cell.id = samples[i])
}

#Finally, remove the raw data to save memory 
rm(data.10x)
```
```{r Merging Seurat Objects,warning=FALSE,message=FALSE,echo=FALSE,include = FALSE}
#Merge the Seurat objects into a single object if there are multiple samples
#We will call this object scrna. Then clean up some space by removing scrna.list.
#Finally, save the merged object as an RDS file. Should you need to load this file into R at any time, it can be done using the readRDS command.

if((length(scrna.list)>1))
{scrna = Reduce(merge, scrna.list)
}else
{scrna=(unlist(scrna.list[[1]]))}

#Finally,remove individual seurat objects to save memory 
rm(scrna.list)
## examine the structure of the Seurat object meta data
str(scrna@meta.data)
saveRDS(scrna, file = sprintf("%s/MergedSeuratObject.rds", scrna_seurat_qc_outdir))
```
#### Calculate some additional metrics for QC: 

*   **Number of genes detected per UMI**: This metric with give us an idea of the complexity of our dataset (more genes detected per UMI, more complex                                           our data). \
*   **Mitochondrial ratio**: This metric will give us a percentage of cell reads originating from the mitochondrial genes. \



```{r Calculate QC metrics,warning=FALSE,message=FALSE,echo=FALSE,include = FALSE}
# Add number of genes per UMI for each cell to metadata
scrna$log10GenesPerUMI <- log10(scrna$nFeature_RNA) / log10(scrna$nCount_RNA)
#Seurat has a convenient function that allows us to calculate the proportion of transcripts mapping to mitochondrial genes. 
#The PercentageFeatureSet() will take a pattern and search the gene identifiers. 
#For each column (cell) it will take the sum of the counts slot for features belonging to the set, 
#divide by the column sum for all features and multiply by 100. 
#Since we want the ratio value for plotting, we will reverse that step by then dividing by 100.
#NOTE: The pattern provided (“^MT-“) works for human gene names. You may need to adjust depending on your organism of interest. 

# Compute percent mito ratio
scrna$mitoPercent <- PercentageFeatureSet(object = scrna, pattern = "^MT-")
scrna$mitoRatio <- scrna@meta.data$mitoPercent / 100
```


#### Visualize QC metrics as a violin plot \


```{r Violin Plot, fig.align = 'center', echo=FALSE}
#VlnPlot(scrna, features = c("nFeature_RNA", "nCount_RNA", "mitoPercent"), ncol = 3)
VlnPlot(scrna, features = c("nFeature_RNA"), ncol = 1)
VlnPlot(scrna, features = c("nCount_RNA"), ncol = 1)
VlnPlot(scrna, features = c("mitoPercent"), ncol = 1)
```
```{r SCRNA Metadata,warning=FALSE,message=FALSE,echo=FALSE}
# Create metadata dataframe
scrna_metadata <- scrna@meta.data

# Add cell IDs to metadata
scrna_metadata$cells <- rownames(scrna_metadata)

# Rename columns
scrna_metadata <- scrna_metadata %>%
  dplyr::rename(seq_folder = orig.ident,
                nUMI = nCount_RNA,
                nGene = nFeature_RNA)
# Create sample column
scrna_metadata$sample <- NA
scrna_metadata$sample <- scrna_metadata$seq_folder


# Add metadata back to Seurat object
scrna@meta.data <- scrna_metadata
saveRDS(scrna, file = sprintf("%s/MergedSeuratObject.rds", scrna_seurat_qc_outdir))
```

#### Cell counts \

```{r Cell count QC,fig.align = 'center',warning=FALSE,message=FALSE,echo=FALSE}
#Assessing the quality metrics
#Cell counts
# Visualize the number of cell counts per sample
scrna_metadata %>% 
  ggplot(aes(x=sample, fill=sample)) + 
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("NCells")
```

#### UMI counts (transcripts) per cell \

*  The UMI counts per cell should generally be **above 500**, that is the low end of what we expect.\
*  If UMI counts are between 500-1000 counts, it is usable but the cells probably should have been sequenced more deeply.\

```{r UMI counts (transcripts) per cell,fig.align = 'center', warning=FALSE,message=FALSE,echo=FALSE}
scrna_metadata %>% 
  ggplot(aes(color=sample, x=nUMI, fill= sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("Cell density") +
  geom_vline(xintercept = 500)
```

#### Genes detected per cell \

*   We have similar expectations for gene detection as for UMI detection, although it may be a bit lower than UMIs. \
*   For high quality data, the proportional histogram should contain a single large peak that represents cells that were encapsulated. \
*   If we see a small shoulder to the right of the major peak or a bimodal distribution of the cells,that can indicate a couple of things. \
        - It might be that there are a set of cells that failed for some reason. \
        - It could also be that there are biologically different types of cells (i.e. quiescent cell populations, less complex cells of interest),              and/or one type is much smaller than the other (i.e. cells with high counts may be cells that are larger in size).  \
*   Therefore, this threshold should be assessed with other metrics.  \

```{r Genes detected per cell,fig.align = 'center',warning=FALSE,message=FALSE,echo=FALSE}
# Visualize the distribution of genes detected per cell via histogram
scrna_metadata %>% 
  ggplot(aes(color=sample, x=nGene, fill= sample)) + 
  geom_density(alpha = 0.2) + 
  theme_classic() +
  scale_x_log10() + 
  geom_vline(xintercept = 300)
# Visualize the distribution of genes detected per cell via boxplot
scrna_metadata %>% 
  ggplot(aes(x=sample, y=log10(nGene), fill=sample)) + 
  geom_boxplot() + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("NCells vs NGenes")
```

#### UMIs vs. genes detected \

*   Two metrics that are often evaluated together are the number of UMIs and the number of genes detected per cell. 
*   Here, we have plotted the number of genes versus the number of UMIs colored by the fraction of mitochondrial reads. \
*   Mitochondrial read fractions are only high in particularly low count cells with few detected genes. \
*   This could be indicative of damaged/dying cells whose cytoplasmic mRNA has leaked out through a broken membrane, and thus, only mRNA located in       the mitochondria is still conserved. These cells are filtered out by our count and gene number thresholds. \
*   Jointly visualizing the count and gene thresholds shows the joint filtering effect. \
*   Cells that are poor quality are likely to have low genes and UMIs per cell, and correspond to the data points in the bottom left quadrant of the      plot. Good cells will generally exhibit both higher number of genes per cell and higher numbers of UMIs. \
*   With this plot we also evaluate the slope of the line, and any scatter of data points in the bottom right hand quadrant of the plot. \
*   These cells have a high number of UMIs but only a few number of genes. \
*   These could be dying cells, but also could represent a population of a low complexity cell type (i.e red blood cells). \

```{r UMIs vs genes detected,fig.align = 'center',warning=FALSE,message=FALSE,echo=FALSE}
# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
scrna_metadata %>% 
  ggplot(aes(x=nUMI, y=nGene, color=mitoRatio)) + 
  geom_point() + 
  scale_colour_gradient(low = "gray90", high = "black") +
  stat_smooth(method=lm) +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = 500) +
  geom_hline(yintercept = 250) +
  facet_wrap(~sample)
```

#### Mitochondrial counts ratio \
* This metric can identify whether there is a large amount of mitochondrial contamination from dead or dying cells. \
* We define poor quality samples for mitochondrial counts as cells which surpass the **0.2** mitochondrial ratio mark, unless of course you are         expecting   this in your sample. \

```{r Mitochondrial counts ratio,fig.align = 'center',warning=FALSE,message=FALSE,echo=FALSE}
# Visualize the distribution of mitochondrial gene expression detected per cell
scrna_metadata %>% 
  ggplot(aes(color=sample, x=mitoRatio, fill=sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  geom_vline(xintercept = 0.2)
```

#### Complexity \
* We can see the samples where we sequenced each cell less have a higher overall complexity, that is because we have not started saturating the         sequencing for any given gene for these samples.  \
* Outlier cells in these samples might be cells that have a less complex RNA species than other cells. \
* Sometimes we can detect contamination with low complexity cell types like red blood cells via this metric. \
* Generally, we expect the novelty score to be **above 0.80**.  \
```{r Complexity,warning=FALSE,fig.align = 'center',message=FALSE,echo=FALSE}
# Visualize the overall complexity of the gene expression by visualizing the genes detected per UMI
scrna_metadata %>%
  ggplot(aes(x=log10GenesPerUMI, color = sample, fill=sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  geom_vline(xintercept = 0.8)
```